---
title: "Post-hocs and planned comparisons"
author: "Your name"
date: "`r date()`"
output: 
  word_document:
    reference_docx: template.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
download.file('https://wkristan.github.io/template.docx', 'template.docx', mode = 'wb')
```

## Import and summary of data

We will work eel larval development data. Import the "eels.xlsx" into a data frame called "eels".

```{r eels.import.data}



```


To plot the means and 95% confidence intervals use summarySE() to get the summary statistics you need:

```{r summarize.data}



```

Then use the summary data set you created to plot the means and 95% confidence intervals using ggplot2:

```{r plot.means.ci}



```

**Question: the stages are in their proper order on the plot. Stage is clearly an ordinal variable, but does that mean that R is considering Stage to be ordinal?**

> 

## Fitting the linear model

Now run the lm() command to fit the model, using HA as the response variable and Stage as the predictor:

```{r fit.lm}



```

**Question: is the Stage term significant? How do you know?**

> 

**Question: can you tell at this point which stages differ?**

> 

Load the emmeans library:

```{r load.emmeans}



```

Run the emmeans() command, from the emmeans library, for the Tukey procedure:

```{r tukey.emmeans}



```

**Question: which stages are different from one another?**

> 

Dunnett's comparisons are all groups against one. Usually the one that is compared against is a "control" group, and all the others are treatments of various kinds. We don't have data like that, but since Stage 1 is set as the baseline group we can use Dunnett's test to compare all of the other stages against Stage 1:

```{r dunnett.glht}



```

**Question: are all of the stages different from Stage 1?**

> 

## Post-hoc comparisons of your choice

We need a matrix that defines comparisons between sequential stages.

```{r seq.comparisons}



```

Now get the multiple comparisons for this set of sequential comparisons of means:

```{r get.seq.comparisons}



```

**Question: which adjacent means are different?**

>

## Orthogonal contrasts

We will use the second set of possible orthogonal contrasts, which compares each stage against the mean of those that follow.

First, make four vectors that have the weights shown in each column of the Set 2 contrast matrix:

```{r make.contrast.vectors}



```

Bind the four vectors of contrast weights into a single contrast matrix:

```{r make.contrast.matrix}



```

Make a copy of the Stage variable, and call it Stage.custom so that we can assign these custom contrasts to it:

```{r make.copy.of.stage}



```

Assign the hand-made custom contrasts to the Stage variable in eels:

```{r assign.contrasts.to.stage}



```

Now we can fit a model using Stage.custom as a predictor and HA as a response, just as we did before, but the coefficient tests will be tests of these four contrasts:

```{r custom.contrasts.test}



```

## Orthogonal polynomials - analyzing ordinal trends

R uses polynomial contrasts by default for ordered factors. If we make a version of the Stage variable that is an ordered factor, and then use it in a linear model, R will give us tests of the ordinal trends across the five stages. 

Make a copy of Stage as an ordered factor:

```{r stage.ordered}



```


To see what the polynomial weights look like, make a set of orthogonal polynomial weights and display them:

```{r orthog.poly.weights}



```

**Question: these weights describe a pattern of change from first to last stage. How? Use Linear and Quadratic weights as an example.**

> 

Fit a model using the ordered factor version of stage:

```{r lm.with.polynomial.contrasts}



```

Get the ANOVA table - it doesn't look any different:

```{r anova.poly.weights}



```

Run Tukey tests on this ordinal stage variable:

```{r tukey.poly.weights}



```

**Question: are there any changes to these Tukey tests compared with the analysis with the unordered factor you did previously?**

> 


The tests of polynomial trends are seen in the coefficient tests:

```{r coefficient.tests.poly.weights}



```

**Question: which polynomial trends were statistically significant?**

> 

## Corrections for multiple p-values

We just completed an analysis of HA, but the data set includes measures of change in water and in neutral sugar as well. Fit a model using water as a response, and Stage as a predictor, then get the summary output:

```{r water.lm}



```
Fit a model using NS as a response, and get the summary output:

```{r ns.lm}



```
**Question: according to the model-level omnibus test (i.e. the p-value at the bottom of the summary output), do NS and water change during development? How do you know?**

>

Calculate the adjusted alpha levels:

```{r adjusted.alphas}



```
**Question: are any of the variables no longer statistically significant at either of these adjusted alpha levels?**

>

## OPTIONAL for the adventurous

Three of the trends were statistically significant. To understand better how the polynomial coefficients work, and what patterns are being tested by each of the trends, we can predict the stage means using the linear trend, then add the quadratic trend, and then the cubic. You'll see that we can only approximately predict the means of stages with the linear trend, but the prediction gets better each time we add another one. Just like we needed four dummy coded columns to reproduce a factor with five stages, we can reproduce the five stage means if we included all four polynomial weights, but the fourth order trend was not significant. We will see why this was the case.

First, get the coefficients:

```{r get.coef}



```

Get the contrast weights:

```{r contrast.weights}



```

Calculate each of the three significant trends - linear first, then add the quadratic, then add the cubic:

```{r trends}



```

We need the trends stacked for graphing - make a data frame from the three statistically significant trends, and then stack the data frame:

```{r make.stacked.trends}



```

Add stage names to the data frame, and re-name the columns:

```{r add.stage.names}



```


Plot the observed means and confidence intervals along with the three trends:

```{r plot.trends}



```
