---
title: "Post-hocs and planned comparisons"
author: "KEY"
date: "`r date()`"
output: 
  word_document:
    reference_docx: template.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
download.file('https://wkristan.github.io/template.docx', 'template.docx', mode = 'wb')
```

## Import and summary of data

We will work eel larval development data. Import the "eels.csv" into a data frame called "eels".

```{r eels.import.data}

library(readxl)
eels <- read_excel("eels.xlsx")
eels$Stage <- factor(eels$Stage)

```


To plot the means and 95% confidence intervals use summarySE() to get the summary statistics you need:

```{r summarize.data}

library(Rmisc)
summarySE(eels, "HA","Stage") -> eels.sumstats

```

Then use the summary data set you created to plot the means and 95% confidence intervals using ggplot2:

```{r plot.means.ci}

library(ggplot2)
ggplot(eels.sumstats, aes(x = Stage, y = HA)) + geom_errorbar(aes(ymin = HA - ci, ymax = HA + ci), width = 0.1) + geom_point()

```

#### Question: the stages are in their proper order on the plot. Stage is clearly an ordinal variable, but does that mean that R is considering Stage to be ordinal?

> No, the categories are in alphabetical order, but the factor is not considered ordered at this point. The ordering will not be used in this first analysis.

## Fitting the linear model

Now run the lm() command to fit the model, using HA as the response variable and Stage as the predictor:

```{r fit.lm}

eels.ha.lm <- lm(HA ~ Stage, data = eels)
anova(eels.ha.lm)

```

#### Question: is the Stage term significant? How do you know?

> Yes, because the p-value for Stage is significant.

#### Question: can you tell at this point which stages differ?

> No, only that there are at least two Stages have mean HA that differ.

Load the emmeans library:

```{r load.emmeans}

library(emmeans)

```

## Tukey comparisons

Run the emmeans() command, from the emmeans library, for the Tukey procedure:

```{r tukey.emmeans}

emmeans(eels.ha.lm, tukey ~ Stage) -> eels.ha.emm
eels.ha.emm

multcomp::cld(eels.ha.emm)

```

#### Question: which stages are different from one another?

> Working from first to last stage, Stage 1 is different from Stage 4 and 5, Stage 2 is different from Stage 3, 4, and 5. Stage 3 is different from Stage 5. STage 4 is different from Stage 5.

## Dunnett's comparisons

Dunnett's comparisons are all groups against one. Usually the one that is compared against is a "control" group, and all the others are treatments of various kinds. We don't have data like that, but since Stage 1 is set as the baseline group we can use Dunnett's test to compare all of the other stages against Stage 1:

```{r dunnett.emmeans}

emmeans(eels.ha.lm, dunnett ~ Stage)

```

#### Question: at what stage did HA become different from Stage 1?

> No, Stage 2 is not, but the rest of the stages are different from Stage 1.

## Sequential comparisons

Comparing each level to the previous one is done with the consec comparisons in emmeans:

```{r consec.emmeans}

eels.ha.consec <- emmeans(eels.ha.lm, consec ~ Stage)
eels.ha.consec

```

#### Question: which stages are different from the previous one?

>

## Orthogonal polynomials - analyzing ordinal trends

R uses polynomial contrasts by default for ordered factors. If we make a version of the Stage variable that is an ordered factor, and then use it in a linear model, R will give us tests of the ordinal trends across the five stages. 

Make a copy of Stage as an ordered factor:

```{r stage.ordered}

eels$Stage.ordered <- factor(eels$Stage, ordered = T)

```

#### Question: refer to the polynomial weights in the instructions. These weights describe a pattern of change from first to last stage. How? Use Linear and Quadratic weights as an example.

> Linear weights start at a negative value and increase by an equal amount across the levels. The quadratic term is a parabola that opens upward, so it starts at positive values, declines to a minimum, and then increases back to the same value as it started.

Fit a model using the ordered factor version of stage:

```{r lm.with.polynomial.contrasts}

eels.op.lm <- lm(HA ~ Stage.ordered, data = eels)
summary(eels.op.lm)

```

Get the ANOVA table - it doesn't look any different:

```{r anova.poly.weights}

anova(eels.op.lm)

```

Run Tukey tests on this ordinal stage variable:

```{r tukey.poly.weights}

emmeans(eels.op.lm, tukey ~ Stage.ordered)

```

#### Question: are there any changes to these Tukey tests compared with the analysis with the unordered factor you did previously?

> No, the comparisons of means are all the same.


The tests of polynomial trends are seen in the coefficient tests:

```{r coefficient.tests.poly.weights}

summary(eels.op.lm)

```

#### Question: which polynomial trends were statistically significant?

> Linear, Quadratic, and Cubic trends were significant.


## Corrections for multiple p-values

We just completed an analysis of HA, but the data set includes measures of change in water and in neutral sugar as well. Fit a model using water as a response, and Stage as a predictor, then get the summary output:

```{r water.lm}

lm(water ~ Stage, data = eels) -> water.lm
summary(water.lm)

```

Fit a model using NS as a response, and get the summary output:

```{r ns.lm}

lm(NS ~ Stage, data = eels) -> ns.lm
summary(ns.lm)

```
#### Question: according to the model-level omnibus test (i.e. the p-value at the bottom of the summary output), do NS and water change during development? How do you know?

> Yes they do, both have p-values less than 0.05.

Calculate the adjusted alpha levels:

```{r adjusted.alphas}

0.05/3
1 - (1 - 0.05)^(1/3)

```
#### Question: are any of the variables no longer statistically significant at either of these adjusted alpha levels?

> All are still statistically significant after adjustment, but water is much closer to the adjusted alpha level.

## Optional: post-hoc comparisons of your choice, and orthogonal contrasts

We need a matrix that defines comparisons between sequential stages.

```{r group.vectors}

s1 <- c(1,0,0,0,0)
s2 <- c(0,1,0,0,0)
s3 <- c(0,0,1,0,0)
s4 <- c(0,0,0,1,0)
s5 <- c(0,0,0,0,1)

```

Define the contrasts you wish to make:

```{r define.contrasts}

list(s1s2.vs.s4s5 = (s1+s2)/2 - (s4+s5)/2, s1.vs.s2s3s4s5 = s1 - (s2+s3+s4+s5)/4) -> custom.contrasts

```


Now get the multiple comparisons for this set of custom contrasts:

```{r test.custom.contrasts}

emmeans(eels.ha.lm, ~Stage) -> eels.ha.emm

contrast(eels.ha.emm, custom.contrasts, adjust = "mvt")

```

#### Question: which comparisons are significant?

> Both are significant.

#### Question: why was it necessary to adjust for multiple comparisons with these contrasts?

> These comparisons are not independent.

## Orthogonal contrasts

We will use the fourth set of possible orthogonal contrasts, which compares each stage against the mean of those that follow.

```{r make.contrast.vectors}

Contrast.1 <- s1 - (s2+s3+s4+s5)/4
Contrast.2 <- (s2+s3)/2 - (s4+s5)/2
Contrast.3 <- (s2+s4)/2 - (s3+s5)/2
Contrast.4 <- (s3+s4)/2 - (s2+s5)/2

```

Make a list of each of the four independent contrasts:

```{r make.contrast.list}

list(s1.vs.s2s3s4s5 = Contrast.1, s2s3.vs.s4s5 = Contrast.2, s2s4.vs.s3s5 = Contrast.3, s3s4.vs.s2s5 = Contrast.4) -> orthog.contrasts

```

Test these contrasts for significance:

```{r test.orthog.contrasts}

contrast(eels.ha.emm, orthog.contrasts)

```

#### Question: which comparisons were significant?

> All of the contrasts except for the last one, which compared the mean of Stage 3 and Stage 4 to the mean of Stage 2 and Stage 5, were significant.

**Question: why was it not necessary to adjust for multiple comparisons, even though there are four contrasts being tested?**

> The contrasts are independent this time, so they do not need to be corrected for multiple comparisons.
