---
title: "Post-hocs and planned comparisons"
author: "Your Name"
date: "`r date()`"
output: 
  word_document:
    reference_docx: template.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
download.file('https://wkristan.github.io/template.docx', 'template.docx', mode = 'wb')
```

## Import and summary of data

We will work eel larval development data. Import the "eels.csv" into a data frame called "eels".

```{r eels.import.data}



```


To plot the means and 95% confidence intervals use summarySE() to get the summary statistics you need:

```{r summarize.data}



```

Then use the summary data set you created to plot the means and 95% confidence intervals using ggplot2:

```{r plot.means.ci}



```

#### Question: the stages are in their proper order on the plot. Stage is clearly an ordinal variable, but does that mean that R is considering Stage to be ordinal?

> 

## Fitting the linear model

Now run the lm() command to fit the model, using HA as the response variable and Stage as the predictor:

```{r fit.lm}



```

#### Question: is the Stage term significant? How do you know?

> 

#### Question: can you tell at this point which stages differ?

> 

Load the emmeans library:

```{r load.emmeans}



```

## Tukey comparisons

Run the emmeans() command, from the emmeans library, for the Tukey procedure:

```{r tukey.emmeans}



```

#### Question: which stages are different from one another?

> 

## Dunnett's comparisons

Dunnett's comparisons are all groups against one. Usually the one that is compared against is a "control" group, and all the others are treatments of various kinds. We don't have data like that, but since Stage 1 is set as the baseline group we can use Dunnett's test to compare all of the other stages against Stage 1:

```{r dunnett.emmeans}



```

#### Question: at what stage did HA become different from Stage 1?

> 

## Sequential comparisons

Comparing each level to the previous one is done with the consec comparisons in emmeans:

```{r consec.emmeans}



```

#### Question: which stages are different from the previous one?

>

## Orthogonal polynomials - analyzing ordinal trends

R uses polynomial contrasts by default for ordered factors. If we make a version of the Stage variable that is an ordered factor, and then use it in a linear model, R will give us tests of the ordinal trends across the five stages. 

Make a copy of Stage as an ordered factor:

```{r stage.ordered}



```

#### Question: refer to the polynomial weights in the instructions. These weights describe a pattern of change from first to last stage. How? Use Linear and Quadratic weights as an example.

> 

Fit a model using the ordered factor version of stage:

```{r lm.with.polynomial.contrasts}



```

Get the ANOVA table - it doesn't look any different:

```{r anova.poly.weights}



```

Run Tukey tests on this ordinal stage variable:

```{r tukey.poly.weights}



```

#### Question: are there any changes to these Tukey tests compared with the analysis with the unordered factor you did previously?

> 


The tests of polynomial trends are seen in the coefficient tests:

```{r coefficient.tests.poly.weights}



```

#### Question: which polynomial trends were statistically significant?

> 


## Corrections for multiple p-values

We just completed an analysis of HA, but the data set includes measures of change in water and in neutral sugar as well. Fit a model using water as a response, and Stage as a predictor, then get the summary output:

```{r water.lm}



```

Fit a model using NS as a response, and get the summary output:

```{r ns.lm}



```

#### Question: according to the model-level omnibus test (i.e. the p-value at the bottom of the summary output), do NS and water change during development? How do you know?

> 

Calculate the adjusted alpha levels:

```{r adjusted.alphas}



```

#### Question: are any of the variables no longer statistically significant at either of these adjusted alpha levels?

> 

## Optional: post-hoc comparisons of your choice, and orthogonal contrasts

We need a matrix that defines comparisons between sequential stages.

```{r group.vectors}



```

Define the contrasts you wish to make:

```{r define.contrasts}



```


Now get the multiple comparisons for this set of custom contrasts:

```{r test.custom.contrasts}



```

#### Question: which comparisons are significant?

> 

#### Question: why was it necessary to adjust for multiple comparisons with these contrasts?

> 

## Orthogonal contrasts

We will use the fourth set of possible orthogonal contrasts.

```{r make.contrast.vectors}



```

Make a list of each of the four independent contrasts:

```{r make.contrast.list}



```

Test these contrasts for significance:

```{r test.orthog.contrasts}



```

#### Question: which comparisons were significant?

>

#### Question: why was it not necessary to adjust for multiple comparisons, even though there are four contrasts being tested?

>
